When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-25 doc/paper/vpd_et_paper_PG_v1.tex
what I sent to Pierre

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity \citep[gentine et al. GRL 2016][]{Berg_2016}.  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature ($\frac{\partial \; e_s}{\partial \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with empirical approaches, which describes the dependence of $g_s$ to VPD. The result for leaf-scale stomatal conductance was:

@@

While Equation 3 can be used in PM (equation 1), it will make analytical work with the function intractable because $A$, net CO$_2$ assimilation, is functionally related to ET itself. To remove the dependence of ET on $A$ we can use the semi-empirical results of \citet{Zhou_2015}. \citet{Zhou_2015} showed that the underlying Water Use Efficiency $uWUE$:

@@

and thus we can evaluate departure from our theory (as a departure of $\sigma$ from unity) The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as the FLUXNET observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have provided an analytical framework for ecosystem response to atmospheric demand perturbations. There are a few subtleties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH} \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be approximately orthogonal.

We note one final comment on our derivation which is relevant for future projects and analysis. If we approximate $c_s$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015, Lin_2015}. This makes Equation \ref{et} a useful alternative to PET in drought indices and hydrometeorological analysis for vegetated surfaces. Equation \ref{et} better reflects the physics of water exchange at the land surface.

@@

Before calculating the sensitivity of ET to VPD, we will consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term, which modifies the magnitude but not the sign of the derivative $\partial ET/\partial VPD$:

@@

and a ``sign'' term, which determines whether the derivative is positive i.e. atmospheric demand driven or negative i.e. physiologically controlled:

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will determine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates and ET decreases with increasing VPD. However, if the first term is larger then atmospheric demand dominates and ET increases with increasing VPD. Given this sign

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions \citep{Zhou_2016, Lin_2015}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Values of $VPD_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems can regulate and mitigate evaporative losses up to a VPD limit, above which atmospheric demand is just too high to be entirely compensated by stomatal regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes. Even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD and ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport and thus carbon allocation (MENCUCCINI) \explain{Not sure citation  - Comparative criteria for models of the vasuclar transport systems of tall trees}.

\begin{table}
  \label{vpd_crit}
\caption{Values of $VPD_{crit}$, where $\frac{\partial \; ET}{\partial \; VPD} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided CITE TABLE. For values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive. \textbf{**** this statistics are dated, need to update***}}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ &  uWUE    & \textbf{$VPD_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

The slope parameter ($g_1$) is additionally a primary determinant of VPD  dependence for the sign term shown in Figure \ref{term3}. Plants that are less willing to trade water for new growth (small $g_1$) will have strong tendency to reduce ET with increasing VPD, and they will be very effective in reducing ET especially at low VPD.  However, at very high VPD gradients in vapor pressure at the leaf scale will begin to dominate leaf-scale plant strategies of stomatal closure (parameterized with $g_1$), so ET response will begin to asymptote towards ecosystem-scale values irregardless of leaf-scale strategy and stomatal closure (through $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET with increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence.  ENF are less willing to trade water for new growth at the leaf scale, so the difference between the response at low VPD when leaf response dominates and high VPD when vapor pressure gradients dominate is much greater. CRO on the otherhand are more willing to trade water for new growth, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively low.

Figure \ref{term3} show clearly that, according to our theory, for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to reduce ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates and empathizes the water conservative strategy of those plants. For CRO on the otherhand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface response to a change in VPD will always me more negative than if the plants are not included in the physics (PET).

@@

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***I think I'm going to remove the bottom plot: its discussion didn't fit in well with the text, and the box plots test inclusion of uncertainty***}}
\label{term3}
\end{figure}

@@

$\frac{P}{T}$ is an air-density term, which we would expect to vary relatively little compared to aerodynamic conductance and $\Delta$. The psychometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clausius-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductance  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

The control of scaling term variability between PFTS by aerodynamic conductance is confirmed by the data (Figure \ref{scale_vary}). Differences between PFT are almost entirely due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The Scaling term for the tree PFTs (DBF, ENF) is generally more than double the scaling terms for the other PFTs which have lower surface roughness and generally smaller aerodynamic conductance.

@@

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conductance, due to differences in surface roughness, between each PFT. In contrast, variability within PFT is also controlled by temperature, through Clausius-Clapeyron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

@@

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
Previous sections only considered idealized scenarios without uncertainty ($\sigma = 1.0$). In this section we will consider bulk statistics, as a function of PFT, of ET response when $\sigma$ is allowed to vary and eddy-covariance derived uncertainty is included (Section \ref{methods}).  These bulk statistics of ET response confirm our expectations based on the theory presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}. For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally much larger magnitude than positive responses (Figure \ref{sign}).

@@

In general the bulk statistics match our expectations based on previous idealized and theoretical analysis, with the possible exception that inclusion of uncertainty shifts grass towards a slightly more negative response than the theory and Figure \ref{sign} would suggest. The bulk statistics motivate a more thorough examination of the structure of uncertainty and more sophisticated validation of our theory's performance against observations.

\begin{table}
  \label{stats}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; VPD}$ as a function of PFT. \textbf{***dated, need to update these statistics***}}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)$ & $\frac{\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{env}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 37.05 & 0.41 & 0.473311\\
CSH & -0.108234 & -0.091526 & 101.72 & 0.88 & 0.931660\\
DBF & -0.012727 & 0.013794 & 39.47 & 0.33 & 0.461674\\
ENF & -0.034087 & 0.000706 & 33.22 & 0.30 & 0.534425\\
GRA & -0.019637 & -0.000921 & 33.60 & 0.35 & 0.631735\\
\hline

\end{tabular}
\end{table}

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now turn to an direct evaluation of the functional forms presented in Section \ref{sign_term} against

t observations taken from a wide range of eddy-covariance datasets (Section \ref{methods}). Figure \ref{agu_fig} shows the distribution of of the sign term (when uncertainty is included), as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a large amount of noise when reproducing the derivatives of ET.

@@

While CSH has a much different functional form of the sign term than DBF, CSH observations also match our theory to leading order. Again, the VPD dependence mostly determined from the slope parameter ($g_1$) closely matches the medians in the observation bins. The VPD-independent, strongly negative response, which is determined by $uWUE$, is also captured. For CSH, there is extremely rare occurrence of positive ET response with VPD, even with uncertainty, so the sign of the theory almost always matches the sign of the observations, even with all of the uncertainty.

As Section \ref{stats_sec} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards positive values relative to the observations, due to some error in $uWUE$. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias the theory explains  the shift towards negative ET response observed in the bulk statistics of Section \ref{stats_sec}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of variability (particularly between sites) for CRO. We do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis, or differences in irrigation practices between the sites and/or years. These deficiencies would explain the inability of our theory to match observations. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{agu_fig}.

@@

The comparison between the theory and observations for ENF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not well captured by observations. However, the theory closely captures the x-intercept and VPD$_{crit}$. So, the theory still captures when the ET response will be positive or negative, but is less accurate with the magnitude of the response. Therefore the theory is still useful for addressing when VPD drives or reduces ET. Also, it's with noting that we used published constants for $g_1$ \citep[from ][]{Lin_2015}, which determines the VPD dependence. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori physical constants whenever possible.

>>>>

The ET response to changes in VPD alters water partitioning between the soil and atmosphere. If plant response reduces ET in response to atmospheric drying then soil moisture will be better conserved. This would seem a sensible evolutionary strategy to cope with aridity. If stomata were fully passive \citep [similar to soil pores, e.g. ][]{Or_2013}, increased atmospheric aridity would further reduce soil moisture. In turn, this would further increase aridity as low soil moisture levels increase the Bowen ratio, and in turn increase temperature and reduce atmospheric humidity \citep[gentine et al. GRL 2016][]{Berg_2016}\explain{Pierre, which publication were you thinking, your Evaporation from Weather Data one?}.  This however would not seem to be a sensible strategy for plants from an evolutionary standpoint.

@@

\section{Methods}
\label{methods}
The Penman-Monteith equation \citep [hereafter PM,][]{Monteith_1965} estimates ET as a function of observable atmospheric variables and surface conductances:
\begin{linenomath*}
  \begin{equation}
      ET = \frac{\Delta R + g_a \rho_a c_p VPD}{\Delta + \gamma(1 + \frac{g_a}{g_s})},
  \end{equation}
\end{linenomath*}
where $\Delta$ is the change in saturation vapor pressure with temperature, given by Clausius-Clapeyron ($\frac{d \; e_s}{d \; T}$), $R$ is the net radiation minus ground heat flux, $g_a$ is aerodynamic conductance, $\rho_a$ is air density, $c_p$ is specific heat of air at constant pressure, $\gamma$ is the psychometric constant, and $g_s$ is the stomatal conductance (Table 1). \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining optimal photosynthesis theory \citep{Farquhar_1980, Katul_2010} with empirical approaches, which describes the dependence of $g_s$ to VPD. The result for leaf-scale stomatal conductance was:

@@

While Equation 3 can be used in PM (Equation 1), it will make analytical work with the function intractable because $A$, net CO$_2$ assimilation, is functionally related to ET itself. To remove the dependence of ET on $A$ we can use the semi-empirical results of \citet{Zhou_2015}, who showed that the underlying Water Use Efficiency $uWUE$:

@@

and thus we can evaluate departure from our theory (as a departure of $\sigma$ from unity). The variability of $\sigma$ across sites and time provides a measure of uncertainty in our model, assumptions, as well as the FLUXNET observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have provided an analytical framework for ecosystem response to atmospheric demand perturbations. There are a few subtleties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH} \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be approximately orthogonal, which is supported by the data.

We note one final comment on our derivation which is relevant for future projects and analysis. If we approximate $c_s$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015} and \citet{Lin_2015}. This makes Equation \ref{et} a useful alternative to PET in drought indices and hydrometeorological analysis for vegetated surfaces. Equation \ref{et} better reflects the physics of water exchange at the land surface.

@@

Before calculating the sensitivity of ET to VPD, we will consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term, which modifies the magnitude but not the sign of the ET response to VPD  ($\frac{\partial \; ET}{\partial \; VPD}$):

@@

and a ``sign'' term, which determines whether ET increases or decreases with VPD  (i.e. atmospheric demand driven or physiologically controlled):

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will determine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates and ET decreases with increasing VPD. However, if the first term is larger then atmospheric demand dominates and ET increases with increasing VPD. Given that the sign term directly address the title of this manuscript, we will examine its functional form first.

\subsection{Functional Form of the Sign Term}
\label{sign_term}
First, we explore the variables within the sign term to gain better intuition on the driver of either the increase or reduction of ET with VPD. CO$_2$ concentration ($c_s$) and the psychometric constant ($\gamma$) are relatively constant to a good approximation so that the variability is dominated by $\sigma$ and $VPD$. $uWUE$ could vary with soil moisture but has been shown to be relatively constant except in very dry soil conditions  (ZHOU ET AL, LIN ET AL)\explain{Pierre, which Zhou and Lin citations were you referring to that show that uWUE changes with very dry soil conditions?}. We thus further assume that the soil is not too dry so that $uWUE$ can be approximated as a constant (within PFT). This then means that the sign term only depends on VPD for a given PFT and is approximately just a function of $VPD$. We can further determine a critical threshold separating an increase from decrease in ET, such that $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

Values of $VPD_{crit}$ as a function of PFT are shown in Table \ref{vpd_crit}. For any values of $VPD$ less than $VPD_{crit}$, ET will decrease with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} < 0$), and for values of $VPD$ greater than $VPD_{crit}$, ET will increase with increasing VPD ($\frac{\partial \; ET}{\partial \; VPD} > 0$). In other words, ecosystems can regulate and mitigate evaporative losses up to a VPD limit, above which atmospheric demand is just too high to be entirely compensated by stomatal regulation. We note however that even though ET increases again above the critical threshold, $VPD_{crit}$, ET is still much lower that potential evaporation as stomata are still strongly regulating vapor fluxes. Even in the absence of soil pore evaporation, stomata do not shut down entirely at very high VPD and ET does not go to zero, because stomata are still slightly open to perform some photosynthesis \citep{Ball_1987, Leuning_1990, MEDLYN_2011}. In addition, upward xylem transport is necessary to maintain phloem transport and thus carbon allocation (MENCUCCINI) \explain{Not sure citation  - Comparative criteria for models of the vasuclar transport systems of tall trees?}.

\begin{table}  
\caption{Values of $VPD_{crit}$, where $\frac{\partial \; ET}{\partial \; VPD} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided CITE TABLE. For values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive. \textbf{**** this statistics are dated, need to update***}}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ &  uWUE    & \textbf{$VPD_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\label{vpd_crit}
\end{table}

@@

The slope parameter ($g_1$) is additionally a primary determinant of VPD  dependence for the sign term shown in Figure \ref{term3}. Plants that are less willing to trade water for new growth (small $g_1$) will have strong tendency to reduce ET with increasing VPD, and they will be very effective in reducing ET especially at low VPD.  However, at very high VPD gradients in vapor pressure at the leaf scale will begin to dominate leaf-scale plant strategies of stomatal closure (parameterized with $g_1$), so ET response will begin to asymptote towards ecosystem-scale values irregardless of leaf-scale strategy and stomatal closure (parameterized with $g_1$). Therefore, plants with a low $g_1$ will have the largest VPD dependence of ET response because the difference in ET response at low VPD (leaf stomatal response dominates) and high VPD (VPD gradient dominates) is largest. This is apparent in the strong VPD dependence of ENF, which has the lowest slope parameter ($g_1=74.31$), in Figure \ref{term3}.

To summarize Figure \ref{term3} and Table \ref{vpd_crit}, CROs are the least water conservative and have the strongest overall tendency to increase ET with increasing VPD, while CSH are the most water conservative and have the strongest tendency to decrease ET with increasing VPD.  ENF ($g_1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g_1 = 183.1$) has the smallest VPD dependence.  ENF are less willing to trade water for new growth at the leaf scale, so the difference between the response at low VPD when leaf response dominates and high VPD when vapor pressure gradients dominate is much greater. CRO on the otherhand are more willing to trade water for new growth, so the leaf scale response is much smaller, and the difference in ET response between low VPD and high VPD is relatively small.

Figure \ref{term3} show clearly that, according to our theory, for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. Therefore, plants are able in most atmospheric conditions to reduce ET in response to increased VPD and thus to reduce water loss. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. For CSH, VPD is always less than VPD$_{crit}$ so that the plant response dominates, emphasizing the water conservative strategy of those plants. For CRO on the otherhand, VPD is almost always higher than VPD$_{crit}$, emphasizing that those plants are water intensive and were actually engineered for photosynthesis rather than water saving. For forests and grass, whether plant response or atmospheric demand dominates is more dependent on the environment, with about half of of observed VPD less than VPD$_{crit}$, i.e. in conditions where plant response dominates. It is also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for potential evaporation i.e. atmospheric demand only, emphasizing that there is still a strong regulation of evaporative flux by stomata and though the plant xylem. The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). Plants are always regulating water exchange from the land surface, even when they reach the limits of they ability to do so. Therefore, the actual land surface response to a change in VPD will always me more negative than if the plants are not included in the physics (PET).

@@

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***I think I'm going to remove the bottom plot: its discussion didn't fit in well with the text, and the box plots test inclusion of uncertainty***}}
\label{sign_plot}
\end{figure}

@@

$\frac{P}{T}$ is an air-density term, which we would expect to vary relatively little compared to aerodynamic conductance and Clausius-Clapeyron ($\Delta$). The psychometric constant ($\gamma$) is also relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature, through the Clausius-Clapeyron relationship $\Delta$. This is as expected, given that the aerodynamic conductance represents the efficiency of exchange between the surface and the atmosphere. As aerodynamic conductance  increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

@@

The control of scaling term variability between PFTs by aerodynamic conductance is confirmed by the data (Figure \ref{scale_vary}). Differences between PFT are almost entirely due to differences in aerodynamic conductance, rather than differences in observed temperature ranges. The scaling term for the tree PFTs (DBF, ENF) is generally more than double the scaling terms for the other PFTs which have lower surface roughness and generally smaller aerodynamic conductance.

@@

To summarize, variability between PFTs is mostly controlled by systematic differences in aerodynamic conductance, due to differences in surface roughness between each PFT. In contrast, variability within PFT is also controlled by temperature, through Clausius-Clapeyron. But, aerodynamic conductance variability generally impacts the scaling term more than temperature, even within PFT.

@@

\subsection{Bulk statistics of ET response to VPD}
\label{stats_sec}
Previous sections only considered idealized scenarios without uncertainty ($\sigma = 1.0$). In this section we will consider bulk statistics of ET response, as a function of PFT, when $\sigma$ is allowed to vary and eddy-covariance derived uncertainty is included (Section \ref{methods}).  These bulk statistics of ET response confirm our expectations based on the theory presented in Sections \ref{sign_term}-\ref{scale_term} (Table \ref{stats}). For all PFTs, mean ET response to increasing VPD is negative. However, ET response evaluated at the average of all variables (e.g. $\sigma$, $T$, $c_s$, $VPD$) is only negative for CSH and GRA. This difference in mean ET response as compared to the ET response at mean environmental conditions is due to the non-linear nature of the response, in which negative responses are generally much larger magnitude than positive responses (Figure \ref{sign_plot}).

@@

In general the bulk statistics match our expectations based on previous idealized and theoretical analysis, with the possible exception that inclusion of uncertainty shifts grass towards a slightly more negative response than the theory and Figure \ref{sign_plot} would suggest. The bulk statistics motivate a more thorough examination of the structure of uncertainty and more sophisticated validation of our theory's performance against observations.

\begin{table}
\caption{Statistics of $\frac{\partial \; ET}{\partial \; VPD}$ as a function of PFT. \textbf{***dated, need to update these statistics***}}
\centering
\begin{tabular}{l c c c c c}
  \hline
PFT & $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)$ & $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)$ & $\frac{\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)*\text{std}(VPD)}{ \frac{\partial \; ET}{\partial \; R}\left(\overline{env}\right)*\text{std}(R)}$ & fraction $\frac{\partial \; ET}{\partial \; VPD} < 0.$ \\
  \hline
CRO & 0.000853 & 0.026241 & 37.05 & 0.41 & 0.473311\\
CSH & -0.108234 & -0.091526 & 101.72 & 0.88 & 0.931660\\
DBF & -0.012727 & 0.013794 & 39.47 & 0.33 & 0.461674\\
ENF & -0.034087 & 0.000706 & 33.22 & 0.30 & 0.534425\\
GRA & -0.019637 & -0.000921 & 33.60 & 0.35 & 0.631735\\
\hline
\end{tabular}
  \label{stats}
\end{table}

\subsection{Validation of theory at eddy-covariance sites}
\label{testing}
We now turn to an direct evaluation of the functional forms presented in Section \ref{sign_term} against observations taken from a wide range of eddy-covariance datasets (Section \ref{methods}). Figure \ref{agu_fig} shows the distribution of of the sign term (when uncertainty is included), as compared to what the theory would predict. Our goal was to capture the leading order behavior of the ET dependence on VPD. Given the assumptions we have made, and the uncertainties of flux tower observations themselves, we expect a large amount of noise when reproducing the derivatives of ET.

@@

While CSH has a much different functional form of the sign term than DBF, CSH observations also match our theory to leading order. Again, the VPD dependence mostly determined from the slope parameter ($g_1$) closely matches the medians in the observation bins. The VPD-independent, strongly negative response, which is determined by $uWUE$, is also captured. For CSH, there is extremely rare occurrence of positive ET response with VPD, even with uncertainty, so the sign of the theory almost always matches the sign of the observations.

As Section \ref{stats_sec} suggested, the theory does not match as well for CRO as for DBF and CSH. At low VPD the VPD dependence (due to $g_1$) is similar to the leading order behavior of the observations, but the theory's curve is shifted towards positive values relative to the observations, due to some bias in $uWUE$. Most strikingly, the observations show a tendency to move back towards negative values at high VPD, which given the functional form of VPD dependence, is behavior that is impossible for the theory to capture. This general positive bias of the theory explains the shift towards negative ET response observed in the bulk statistics of Section \ref{stats_sec}. The divergence between the theory and the observations for CRO could have been expected. For practical reasons, our theory does not include important sources of variability (particularly between sites) for CRO. We do not account for seasonal changes in LAI, variability in crop height and surface roughness, differences in C3 vs C4 photosynthesis, or differences in irrigation practices between the sites and/or years. These deficiencies would explain the inability of our theory to match observations. For example, a superposition of sites with C3 plants (at low environmental VPD) and C4 plants (at high environmental VPD) would explain observed shift back towards negative ET response at high VPD when all sites are binned together, as in Figure \ref{agu_fig}.

@@

The comparison between the theory and observations for ENF is unique from the rest of the PFTs. The VPD dependence of the theory, determined by the slope parameter ($g_1$), is not well captured by observations. However, the theory closely captures the x-intercept and VPD$_{crit}$. So, the theory still captures when the ET response will be positive or negative, but is less accurate with the magnitude of the response. Therefore the theory is still useful for addressing when VPD drives or reduces ET. Also, it's with noting that we used published constants for $g_1$ \citep[from ][]{Lin_2015}, which determines the VPD dependence. We could have fit our own constants so that the theory would best match the observations, but our goal was to use a priori results whenever possible.
